A radiocarbon data based evaluation of Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe

### 2019-01-10 article/01_Introduction.Rmd
improvements inn the introduction section

This article employs some of the methods that arose within archaeology's contemplation of *Cultural transmission theory* to trace the development of four major Bronze Age burial traditions in Europe. The cultural-historical evolution of archaeological cultures and ethnic groups during the European Bronze Age is complex, and its reconstruction an ongoing endeavour of archaeology. Within this history of constant emergence, displacement and demise many different forms of burial traditions were practised. Most of them can be distinguished by the criteria *burial construction* and *treatment of the body*. Bronze Age burials are either flat graves or part of a burial mound, and they contain either untreated or cremated human remains. In a nutshell, the classic narrative of the development of these rites in Europe presented e.g. by @hardingEuropeanSocietiesBronze2000a (following @hauslerBestattungssittenFruhenBronzezeit1977; @hauslerGrabUndBestattungssitten1994; @hauslerTotenorientierungUndGeographischer1996) can be summarised as follows: In the early and middle Bronze Age inhumations dominated in different variations. Cremation burials in this time only occur in the Hungarian lowlands. Burial mounds are concentrated in parts of the Balkans as well as in eastern, western and northern Europe, while in central and southern Europe flat graves -- mostly burials -- predominate. In the Middle Bronze Age, barrows gained importance in Western and Central Europe. This observation is highly linked to the emergence of the *Tumulus culture* in Central Europe. In the late Bronze Age, the Urnfield period, cremation became the most common burial custom in the context of *Urnfield culture*.

Harding's narrative is, even when taking into account its enormous simplification, partially misleading. Cremation, for example, was a lot more common already in the early and middle Bronze Age, though not in the context of the later Urnfield culture (e.g. in the Benelux area [@fokkensBronzeAgeLow2013] or Great-Britain [@robertsBritainIrelandBronze2013]). Beyond simple research progress, one additional reason why major misconceptions like this can appear and thrive might be a lack of unified data repositories and universally acknowledged proxies for the distribution of cultural phenomena in European archaeology. Instead of directly querying quantitative data the typical process for the creation of large-scale overviews is to excerpt the literature of all relevant local research traditions. The present article, in contrast, explores whether openly available bulk radiocarbon data could allow reconstructing the trends within the distribution of burial rites. This approach could set a precedent for a more reproducible and statistically sound observation of the main shifts. The quantitative comparison of macroregional trends also paves the way for an analysis of patterns of interaction intensity on a continental scale. A core question is how cultural and spatial distance are related throughout the European Bronze Age.

Already the assumption that the cremation aspect of Early Bronze Age burial rituals in Britain can be directly compared to the same aspect in Hungary requires a theoretical foundation. *Cultural transmission theory* provides a framework where this claim can be explored as a hypothesis for further research. The spread of ideas in human populations over space and time is a major topic of *Cultural evolution theory* [@creanzaCulturalEvolutionaryTheory2017]. This line of thought is at least as old as theories of biological evolution and underwent a long and versatile research history from *Evolutionism* and *Social Darwinism* in the 19th and early 20th century, against *Cultural relativism* and back to *Neoevolutionism* up until the nineteen sixties and seventies [@petermannGeschichteEthnologie2004]. In the late seventies, the discussion took another turn away from the political and social structure to the more general observation of human behaviour. Three major theoretical movements emerged in this context: *Evolutionary psychology*, *Human behavioural ecology* and *Dual inheritance theory* [@smithThreeStylesEvolutionary2000]. The latter -- heavily popularised by Richard Dawkin's *Memetics* [@dawkinsSelfishGene1976] -- establishes the notion of culture acting as secondary heredity system in humans [@henrichDualinheritanceTheoryEvolution2007]. Just like in biological evolution there might be selection and mutation within *Social learning* and therefore methods of *Population genetics* could be employed within cultural anthropology -- of course with the much faster pace of cultural exchange in mind. A lot of theoretical and quantitative research arose from this assumptions: e.g. @cavalli-sforzaCulturalTransmissionEvolution1981, @boydCultureEvolutionaryProcess1985, @teltserEvolutionaryArchaeologyMethodological1995 or @obrienEvolutionaryArchaeologyTheory1996.

>>>>

This article employs some of the methods that arose within archaeology's contemplation of *Cultural transmission theory* to trace the development of four major Bronze Age burial traditions in Europe. The cultural-historical evolution of archaeological cultures and ethnic groups during the European Bronze Age is complex, and its reconstruction an ongoing endeavour of archaeology. Within this history of constant emergence, displacement and demise many different forms of burial traditions were practised. Most of them can be distinguished by the criteria *construction of the grave* and *treatment of the body*. Bronze Age burials are either flat graves or part of a burial mound, and they contain either untreated or cremated human remains.

In a nutshell, the classic narrative of the development of these rites in Europe presented e.g. by @hardingEuropeanSocietiesBronze2000a (following @hauslerBestattungssittenFruhenBronzezeit1977; @hauslerGrabUndBestattungssitten1994; @hauslerTotenorientierungUndGeographischer1996) can be summarised as follows: In the early and middle Bronze Age inhumations dominated in different variations. Cremation burials in this time only occur in the Hungarian lowlands. Burial mounds are concentrated in parts of the Balkans as well as in eastern, western and northern Europe, while in central and southern Europe flat graves -- mostly burials -- predominate. In the Middle Bronze Age, barrows gained importance in Western and Central Europe. This observation is highly linked to the emergence of the *Tumulus culture* in Central Europe. In the late Bronze Age, the Urnfield period, cremation became the most common burial custom in the context of *Urnfield culture*. Harding's narrative is, even when taking into account its enormous simplification, partially misleading. Cremation, for example, was a lot more common already in the early and middle Bronze Age, though not in the context of the later Urnfield culture (e.g. in the Benelux area [@fokkensBronzeAgeLow2013] or Great-Britain [@robertsBritainIrelandBronze2013]).

How can major misconceptions like this arise and persist in European archaeology? Beyond missing data due to preservation and research history, an additional reason might be a lack of unified data repositories and universally acknowledged methods to measure the spatial and temporal distribution of cultural phenomena. Instead of directly querying archives with quantitative information, the typical process for the creation of large-scale syntheses is still to manually excerpt the literature of all relevant, local research traditions. The present article, in contrast, explores the former strategy: Does openly available, bulk radiocarbon data allow reconstructing the main trends within the distribution of burial rites in Bronze Age Europe? This approach could set a precedent for a more reproducible and statistically sound research practice. If so, then the quantitative comparison of macroregional trends could also pave the way for an analysis of interaction intensity on a long-term continental scale: Do time series constructed from radiocarbon data allow meaningful observations concerning cultural and spatial distance?

Already the assumption that the cremation aspect of Early Bronze Age burial rituals in Britain can be directly compared to the same aspect in Hungary requires a theoretical foundation. *Cultural transmission theory* provides a framework where this claim can be explored as a hypothesis for further research. The spread of ideas in human populations over space and time is a major topic of *Cultural evolution theory* [@creanzaCulturalEvolutionaryTheory2017]. This line of thought is at least as old as theories of biological evolution and underwent a long and versatile research history from *Evolutionism* and *Social Darwinism* in the 19th and early 20th century, against *Cultural relativism* and back to *Neoevolutionism* up until the nineteen sixties and seventies [@petermannGeschichteEthnologie2004]. In the late seventies, the discussion took another turn away from the political and social structure to the more general observation of human behaviour. Three major theoretical movements emerged in this context: *Evolutionary psychology*, *Human behavioural ecology* and *Dual inheritance theory* [@smithThreeStylesEvolutionary2000]. The latter -- heavily popularised by Richard Dawkin's *Memetics* [@dawkinsSelfishGene1976] -- establishes the notion of culture acting as secondary heredity system in humans [@henrichDualinheritanceTheoryEvolution2007]. Just like in biological evolution there might be selection and mutation within *Social learning* and therefore methods of *Population genetics* could be employed within cultural anthropology -- of course with the much faster pace of cultural exchange in mind. A lot of theoretical and quantitative research arose from these assumptions: e.g. @cavalli-sforzaCulturalTransmissionEvolution1981, @boydCultureEvolutionaryProcess1985, @teltserEvolutionaryArchaeologyMethodological1995 or @obrienEvolutionaryArchaeologyTheory1996.
